Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-24 paper/paper.tex
Natural transformations

@@

\subsection{Sorting Algorithms}\label{sec:sorting}
We now focus on divide-and-conquer sorting algorithms and demonstrate how to use
program calculation to apply a fusion optimisation. In the supplemental material
accompanying this paper, we formalise both mergesort and quicksort, but in this
section, we focus solely on quicksort.

The recursive structure of quicksort is given by the following
functor
\begin{coqcode}
Inductive ITreeF N X := i_leaf | i_node (n : N) (l r : X)
\end{coqcode}
The idea is that a list is either empty or is split into a pivot which is
represented by node of type \coq{N} and the two sublists of type \coq{X}.

@@

\subsubsection{Fusing a divide-and-conquer computation}
We now show how we can use program calculation techniques to fuse a traversal
with a  divide-and-conquer algorithm to obtain a new program that only performs
recursion once instead of twice. In particular, we can map a function to the
result of quicksort and obtain a new program which orders the elements and
computes the function on every element by traversing the list only once. The
composition of these two programs is given by the following definition:
\begin{coqcode}
Definition qsort_times_two := Lmap times_two \o hylo merge qsplit.
\end{coqcode}
where \coq{Lmap times_two} is a list map function defined as a hylomorphism,
and \coq{times_two} multiplies every element of the list by two.
We can use Coq's generalised rewriting and \coq{hylo_fusion_l} to fuse
\coq{times_two} into \coq{hylo merge qsplit} which gives the program
\coq{hylo (merge \o natural times_two) qsplit}.
In this definition, \coq{natural} defines a natural transformation by
applying \coq{times_two} to the shapes thus multiplying every pivot by two.
%% Our formalisation contains a proof that \coq{natural} is indeed a natural
%% transformation, which relies on the fact that it preserves the structure of the
%% shapes and, therefore, the validity of the positions.
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}
The extracted OCaml code is a single recursive traversal. Note the similarity to
a hand-written implementation, and that it has been derived by fusing two
different recursive programs using regular Coq rewrite tactics.

>>>>

Given two functors $F, G : \CatC \to \CatD$, a \emph{natural transformation} is
a map between functors is a family of morphisms $\phi_{X} : FX \to GX$ indexed
by the objects $X \in \Obj(\CatC)$ and such that it is \emph{natural} in $X$,
that is $ G(f) \cdot \phi_{X} = \phi_{Y} \cdot F(f)$. Intuitively, a natural
transformation transforms the structure of a functor into the structure of
another functor without assuming what is the type of data contained in them, for
example, a program $f : \texttt{List }X \to \texttt{Maybe} X$ which returns
nothing if the list is empty and the head of the list otherwise needs not to
know the type of the data.

@@

\subsection{Sorting Algorithms}\label{sec:sorting}
We now focus on divide-and-conquer sorting algorithms and demonstrate how to use
program calculation to apply a fusion optimisation. In the supplemental material
accompanying this paper, we formalise both mergesort and quicksort, but in this
section, we focus solely on quicksort. The recursive structure of quicksort is
given by the following functor
\begin{coqcode}
Inductive ITreeF N X := i_leaf | i_node (n : N) (l r : X)
\end{coqcode}
The idea is that a list is either empty or is split into a pivot which is
represented by node of type \coq{N} and the two sublists of type \coq{X}.

@@

\subsubsection{Fusing a divide-and-conquer computation}
We now show how we can use program calculation techniques to fuse a traversal
with a  divide-and-conquer algorithm to obtain a new program that only performs
recursion once instead of twice. In particular, we can map a function to the
result of quicksort and obtain a new program which orders the elements and
computes the function on every element by traversing the list only once. The
composition of these two programs is given by the following definition:
\begin{coqcode}
Definition qsort_times_two := Lmap times_two \o hylo merge qsplit.
\end{coqcode}
where \coq{Lmap times_two} is a list map function defined as a hylomorphism, and
\coq{times_two} multiplies every element of the list by two. We can use Coq's
generalised rewriting and \coq{hylo_fusion_l} to fuse \coq{times_two} into
\coq{hylo merge qsplit} which gives the program \coq{hylo (merge \o natural times_two) qsplit}.
In this definition, \coq{natural} defines a natural transformation by applying
\coq{times_two} to the shapes thus multiplying every pivot by two.

%% Our formalisation contains a proof that \coq{natural} is indeed a natural
%% transformation, which relies on the fact that it preserves the structure of the
%% shapes and, therefore, the validity of the positions.
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}
The extracted OCaml code is a single recursive traversal. Note the similarity to
a hand-written implementation, and that it has been derived by fusing two
different recursive programs using regular Coq rewrite tactics.
